High Performance P$^{3}$M N-body code: CUBEP$^3$M

### 2012-08-13 paper_new.tex
latest changes about the new kernels, more details about initial conditions and memory footprint, and included recommendations by Ilian.

\title[{\small}  High Performance P$^{3}$M N-body code: CUBEP$^3$M]{{\small} High Performance P$^{3}$M N-body code: CUBEP$^3$M}
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, J.D. Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$, 
Ilian T. Iliev$^{3}$, Hugh Merz$^{4}$, \newauthor
J.D. Emberson$^{1,5}$ and Vincent Desjacques$^{6}$\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Ontario, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Astronomy Centre, Department of Physics and Astronomy, Pevensey II Building, University of Sussex, BN1 9QH, Brighton, United Kingdom\\
$^{4}$SHARCNET, Laurentian University, P3E 2C6, Ontario, Canada\\
$^{5}$Department of Astronomy and Astrophysics, University of Toronto, M5S 3H4, Ontario, Canada\\
$^{6}$Universit\'{e} de Gen\`{e}ve and Center for Astroparticle Physics, 24 Quai Ernest Ansermet, 1211 Gen\`{e}ve 4, Switzerland}

@@

\begin{abstract}
This paper presents {\small CUBEP$^3$M}, a high performance cosmological N-body code
and describes many utilities and extensions that have been added to the standard package, including a runtime halo finder,
a non-Gaussian initial conditions generator, a tuneable accuracy, and a system of unique particle identification.
{\small CUBEP$^3$M} is fast, has a low memory imprint, and 
has been run on up to $20,000$ cores, achieving close to ideal weak scaling
even at this scale.
It is well suited for a vast number of science applications that 
require either large samples of non-linear realizations or 
very large dark matter N-body simulations,
including  cosmological reionization, baryonic acoustic oscillations, weak lensing or
non-Gaussian statistics.
We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
of the code and its utilities, when applicable.
\end{abstract}

@@

Although such codes are needed to study systems that are spatially heterogeneous like individual clusters or haloes, AGNs or other compact objects,
many applications are interested in studying large cosmological volumes in which the matter distribution is rather homogeneous.
In such environments, the load balancing algorithm can be removed, and the effort
can thereby be put towards speed, memory compactness and scalability.  
{\small PMFAST} (\cite{2005NewA...10..393M}, MPT hereafter) was one of the first code designed specifically to optimize the PM algorithm,
both in terms of speed and memory usage, and uses a two-level mesh algorithm based on the gravity solver of \cite{2003AAS...203.9703T}.
The long range gravitational force is computed on a  grid $4^{3}$ times coarser, such as to minimize the {\small MPI} communication time
and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
to be stored at a given time, allowing for {\small OPENMP} parallel computation.
This setup enables the code to evolve large cosmological systems both rapidly and accurately, even on relatively modest clusters.
One of the main advantages of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
and the global {\small MPI} communications are cut down to the minimum: for passing particles at the beginning of each time step,
and  for computing the long range FFTs.

@@

The paper is structured as follow: section \ref{sec:structure} reviews the structure and flow of the main code;
section \ref{sec:Poisson} describes how Poisson equation is solved on the two-mesh system;
we then present in section \ref{sec:scaling} the scaling of the code to very large problems,
including much larger runs that were produced at various high performance computing centers;
section \ref{sec:accuracy}  discuss the accuracy and systematic effects of the code.
 We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions
 to the default configuration, and conclude afterwards.

@@

When it comes to finding haloes at run time, this buffer can create a problem, because a large object located close to the boundary can have a radius larger than the buffer zone, in which case it would be truncated and be assigned a wrong mass, center of mass, etc. 
It could then be desirable to increase the buffer zone around each tile, at the cost of a loss of memory dedicated to the actual physical volume, and the code is designed to allow for such a flexibility.

@@

Because the coarse grid arrays require $4^3$ times less memory per node, 
it does not contribute much to the total memory requirement, and the bulk of the foot-print is 
concentrated in a handful of fine grid arrays.
Some of these are only required for intermediate steps of the calculations, 
hence it is possible to hide therein many coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\item{{\tt PID} stores the unique particle identification tags.}
\end{enumerate}
The particle ID is a feature that can be switched off by removing a compilation flag, 
and allows to optimize the code for higher resolution configurations.
In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
particle phase space array, in which case it uses about $60$ bytes per particles. 
For a fixed problem size, this can be achieved by increasing the number of tiles, thereby reducing the size of the local fine mesh arrays.
There is a limit at which we can do such a volume breakdown, though, since the physical 
volume on each tile must be much larger than the connection length between the two meshes (see section \ref{sec:Poisson}).
%
%uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
For comparison, (Lean) {\small GADGET-2} code uses around $90$ per particles.
%\subsection{Code overview}
%\label{subsec:overview}

@@

If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and writes them to file. 
These restart files start with a small header that 
contains the local  number of particles, the current redshift and time step, and the constraints on the time step jump
from the previous iteration; the header is followed by the phase space of each particles. The particle identification tags are written 
with a similar fashion in distinct files to simplify the post-processing coding. 
This general I/O strategy allows for highly efficient {\small MPI} parallel read and write,
by default in binary format for compactness, and has been shown to scale well to large data sets.

@@

Other constraints that need to be considered is that any decomposition geometry somehow 
limits permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
--  such as {\small CUBEP$^3$M} -- requires the number of {\small MPI} processes to be a perfect cube.
Also,  since the decomposition is volumetric, as opposed to density dependent -- 
it suffers from load imbalance whenever the particles are not evenly distributed.
However, large scale cosmology problems are very weakly affected by this effect, which
can generally  be overcome by allocating a mild amount of extra memory per node.
As mentioned in section \ref{sec:scaling}, large runs can even be optimized by  allocating only a small amount for the first part of it,
follow by a restart with a configuration that allows more memory per node.

@@

The force of gravity on a mesh can be computed either with a gravitational potential  kernel $\omega_{\phi} ({\bf x})$
  or a force  kernel $\omega_{F} ({\bf x})$.
Gravity fields are curl-free, which allows us to relate the potential $\phi({\bf x})$ to the source term via Poisson's equation: 
\begin{eqnarray}
\nabla^{2}\phi({\bf x}) = 4 \pi G \rho({\bf x})
\label{eq:poisson}
\end{eqnarray}
We solve this equation in Fourier space, where we write:
\begin{eqnarray}
 \tilde{\phi}({\bf k}) = \frac{-4 \pi G \tilde{\rho}({\bf k})}{k^{2}} \equiv \tilde{\omega}_{\phi}({\bf k})\tilde{\rho}({\bf k})
\label{eq:poissonFourier}
\end{eqnarray}
The potential in real space is then obtained with an inverse Fourier transform, and the kernel becomes $\omega_{\phi} ({\bf x}) = -G/r$.
Using the convolution theorem, we can write
\begin{eqnarray}
 \phi({\bf x}) = \int \rho({\bf x'}) \omega_{\phi}({\bf x'} - {\bf x}) d{\bf x'}   
\label{eq:poisson_solution_pot}
\end{eqnarray}
and
\begin{eqnarray}
{\bf F}({\bf x}) = - m {\bf \nabla} \phi({\bf x}) 
\label{eq:Force_sol}
\end{eqnarray}
Although this approach is fast, it involves a finite differentiation at the final step, which enhances the numerical noise.
We therefore opt for a force kernel, which is more accurate, even though it requires four extra Fourier transforms.
In this case, we must solve the convolution in three dimensions and define the force kernel {\boldmath $\omega$}$_{F}$ such as:
\begin{eqnarray}
 {\bf F}({\bf x}) =  \int \rho({\bf x'}) \mbox{\boldmath $\omega$}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
\label{eq:poisson_solution_force}
\end{eqnarray}
Because the gradient acting on \ref{eq:poisson_solution_pot} affects only unprime variables, we can express the force kernel as a gradient of the potential kernel. Namely:  
\begin{eqnarray}
\mbox{\boldmath $\omega$}_{F}({\bf x}) \equiv - {\bf \nabla}\omega_{\phi}({\bf x}) = - \frac{mG \hat{\bf r}}{r^{2}}
\end{eqnarray}

@@

Since these calculations are performed on two grids of different resolution, a sampling window function must be convoluted 
both with the density and the kernel (see [Eq. 7-8] of MPT).
When matching the two force kernels, the long range force is always on the low side close to the cutoff region, whereas the short range force is uniformly scattered across the theoretical $1/r^2$ value -- intrinsic features of the CIC and NGP interpolation schemes respectively.  By performing force measurements on two particles randomly placed in the volume, we identified a small region surrounding the cutoff length in which we empirically adjust both kernels such as to improve the match. Namely, for $14 \le r \le 16$, $\mbox{\boldmath $\omega$}_{s}({r}) \rightarrow 0.985\mbox{\boldmath $\omega$}_{s}({ r})$,
and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow 1.2\mbox{\boldmath $\omega$}_{l}({ r})$.

@@

The choice of softening length is motivated by a trade off between accuracy and run time.
 Larger values reduce the structure formation but run quicker. 
 We show in Fig. \ref{fig:rsoft} the impact of changing this parameters on the power spectrum.
 For this test, we produced SciNet256 simulations, starting at $z=100$ and evolving to $z=0.5$, with a box size of $100 h^{-1}\mbox{Mpc}$.
 We also record in Table \ref{table:rsoft} the real run time for each trials, from where 
we see the strong effect of computation time the softening length has.
In this test, which is purposefully probing rather deep in the non-linear regime, reducing the softening length
to half it value ($1/20^{th}$ of a grid cell) double the run time. The gain in power spectrum is less than two per cent.
Similarly, increasing to $0.2$ the softening length reduces almost by a half the run time, but suffers from a five per cent loss in power at the resolution turn-around.
One tenth of grid cell seems to be the optimal choice in this trade off, and is therefor the default value, however it should really be considered a free parameter.

@@

\begin{table}
\begin{center}
\caption{Scaling in {\small CPU} resources as a function of the softening length.}
\begin{tabular}{|l|c|c|}
\hline 
$r_{soft}$         & time (h)   \\                 
\hline
 $0.5$ & 5.75 \\
 $0.3$ & 8.17\\
 $0.2$ & 10.09 \\
 $0.1$ & 18.67\\
 $0.05$ & 31.07 \\
% $0.03$ & XXX\\
% $0.01$ & XXX\\
\hline
\end{tabular}
\label{table:rsoft}
\end{center}
\end{table}

@@

We have also ran {\small CUBEP$^3$M} on a much larger number of cores, 
from 8000 to up to 21,976, with $5488^3$-$6000^3$ (165 to 216 billion) 
particles on Ranger and on JUROPA at the J\"ulich Supercomputing Centre in Germany, 
which is an Intel Xeon X5570 
(Nehalem-EP) quad-core 2.93 GHz system, also interconnected with Infiniband.
Since it is not practical to perform dedicated scaling tests on such a large number of
computing cores, we instead list in Table \ref{summary_scaling_table2} 
the data directly extracted from production runs. We have found the 
code to scale within 1.5 per cent of ideal up to 4096 cores. 
For larger sizes ($\ge$10,976 cores), the scaling is less ideal,
due to increased communication 
costs, I/O overheads (a single timeslice of $5488^3$ particles is 3.6 TB)
and load balancing issues, but still within $\sim20$ per cent of ideal. 
These first three Ranger runs were performed 
with 4 {\small MPI} processes and 4 threads per Ranger node (`4way')\footnote{For these very large runs, 
we used a NUMA script {\it tacc\_affinity}, specially-provided by the technical staff, 
that bind the memory usage to local sockets, thus ensuring memory affinity. 
This becomes important because the memory sockets per node 
(32 GB RAM/node on Ranger) are actually not equal-access. Generally, the local 
memory of each processor has much shorter access time.}.

Furthermore, due to the increasing clustering of structures at those small 
scales, some of the cuboid sub-domains came to contain a number of particles
well above the 
average, thereby requiring more memory per {\small MPI} process
in order to run until the end. 
As a consequence,  throughout most of their late evolution, the 
largest two of these simulations were run with 4096 and 21,952 cores and 
with only 2 {\small MPI} processes and 8 threads per Ranger node (`2way'), which on 
Ranger allows using up to 16 GB of RAM per {\small MPI} process\footnote{In order to insure 
local memory affinity,  a second special NUMA control script, {\it tacc\_affinity\_2way}, 
was developed for us by the TACC technical staff and allowed to run more efficiently 
in this mode.}. Because each processor accesses memory that is not fully local, this configuration does affect the 
performance somewhat, as does the imperfect load balancing that arises in such situations.
This can be seen in the rightmost point of Fig. \ref{scaling} (right panel), where the scaling is $42$ per cent below the ideal.
We note that we still get $\sim1.5$ speedup
from doubling the core count, even given these issues. Overall the
code scaling performance is thus satisfactory even at extremely large number of cores.
 We expect the code to handle even larger problems efficiently, and is thus well suited to run on
  next generation Petascale systems.

@@

 This section describes the systematics effects that are inherent to our P$^{3}$M algorithm.
 We start with a demonstration of the accuracy with a power spectrum measurement of a RANGER4000 simulation. 
 The halo mass function also assess the code capacity to model gravitational collapse, 
 but depends on the particular halo finder used.
 We thus post-pone the discussion on this aspect until section \ref{sec:halo},
and focus for the moment on the particles only. In addition to the power spectrum, we further quantify the accuracy of the force calculation
 with comparisons to Newton's law of gravitation.

@@

\begin{figure*}%[ht]
  \begin{center}
%    \includegraphics[width=5.2in]{graphs/0.000xz_w_halos_pink_blue_enh-1.eps}
\includegraphics[width=5.2in]{graphs/0.000proj_1000x1000.eps}
  \caption{Detail of a dark matter density projection, measured at $z=0$ in a RANGER4000 simulation.
  Particles are shown in pink, haloes in blue in the online version. The full projection is 64 times larger,
  but does not provide much details on the structure on a paper format.
    \label{fig:density}}
\end{center}
\end{figure*}

@@

The top right panel of Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial direction (top) and the fractional 
tangential contribution (bottom), also calculated from a single time step.
The fractional error of the radial component is up to 60 per cent at sub-grid distances, but remains under ten per cent for  distances larger than a few cells. 
The transverse fractional error peaks exactly at the grid distance,  and is about two time smaller than the radial counterpart.

Although these scatter plots show the contributions from individual particles and cells, it is not clear
whether the mean radial force felt is higher or lower the predictions, hence we rebin the results
in 50 logarithmically spaced bins and compute the mean and standard deviation. 
We show this plot in the middle left panel of Fig. \ref{fig:den_force_fracErr}, where we observe that on average,
there is a loss of force of the order of unity in the range $ 0.4 < r < 1$ (in units of fine grid), but the numerical calculations are otherwise
in excellent agreement with Newton's law.
The transverse force is somehow ill-defined, since it is by definition a positive number.
Since we know that it averages out to zero over many time steps, we plot only the scatter about the mean, represented in the figure by the solid line.

>>>>

\title[CUBEP$^3$M]{High Performance P$^{3}$M N-body code: CUBEP$^3$M}
\author[Joachim Harnois-D\'{e}raps, Ue-Li Pen, Ilian T. Iliev, Hugh Merz, J.D. Emberson, Vincent Desjacques]{Joachim Harnois-D\'{e}raps$^{1,2}$ 
\thanks{E-mail: jharno@cita.utoronto.ca},  Ue-Li Pen$^{1}$, 
Ilian T. Iliev$^{3}$, Hugh Merz$^{4}$, \newauthor
J.D. Emberson$^{1,5}$ and Vincent Desjacques$^{6}$\\
%\footnotemark[1]\thanks{This file has been amended to
%highlight the proper use of \LaTeXe\ code with the class file.
%These changes are for illustrative purposes and do not reflect the
%original paper by A. V. Raveendran.}\\
$^{1}$Canadian Institute for Theoretical Astrophysics, University of
Toronto, M5S 3H8, Ontario, Canada\\
$^{2}$Department of Physics, University of Toronto, M5S 1A7, Ontario,  Canada\\
$^{3}$Astronomy Centre, Department of Physics and Astronomy, Pevensey II Building, University of Sussex, BN1 9QH, Brighton, United Kingdom\\
$^{4}$SHARCNET, Laurentian University, P3E 2C6, Ontario, Canada\\
$^{5}$Department of Astronomy and Astrophysics, University of Toronto, M5S 3H4, Ontario, Canada\\
$^{6}$Universit\'{e} de Gen\`{e}ve and Center for Astroparticle Physics, 24 Quai Ernest Ansermet, 1211 Gen\`{e}ve 4, Switzerland}

@@

\begin{abstract}
This paper presents {\small CUBEP$^3$M}, a high performance, publicly-available, cosmological N-body code
and describes many utilities and extensions that have been added to the standard package, including a runtime halo finder,
a non-Gaussian initial conditions generator, a tuneable accuracy, and a system of unique particle identification.
{\small CUBEP$^3$M} is fast, has a low memory imprint, and 
has been run on up to $20,000$ cores, achieving close to ideal weak scaling
even at this scale.
It is well suited and has already been used for a broad number of science applications that 
require either large samples of non-linear realizations or 
very large dark matter N-body simulations,
including  cosmological reionization, baryonic acoustic oscillations, weak lensing or
non-Gaussian statistics.
We discuss the structure, the accuracy, any known systematic effects, and the scaling performance
of the code and its utilities, when applicable.
\end{abstract}

@@

Although such codes are needed to study systems that are spatially heterogeneous like individual clusters or haloes, AGNs or other compact objects,
many applications are interested in studying large cosmological volumes in which the matter distribution is rather homogeneous.
In such environments, the load balancing algorithm can be removed, and the effort
can thereby be put towards speed, memory compactness and scalability.  
{\small PMFAST} (\citet{2005NewA...10..393M}, MPT hereafter) was one of the first code designed specifically to optimize the PM algorithm,
both in terms of speed and memory usage, and uses a two-level mesh algorithm based on the gravity solver of \citet{2003AAS...203.9703T}.
The long range gravitational force is computed on a  grid $4^{3}$ times coarser, such as to minimize the {\small MPI} communication time
and to fit in system's memory. The short range is computed locally on a finer mesh, and only the local sub-volume needs 
to be stored at a given time, allowing for {\small OPENMP} parallel computation.
This setup enables the code to evolve large cosmological systems both rapidly and accurately, even on relatively modest clusters.
One of the main advantages of {\small PMFAST} over other PM codes is that the number of large arrays is minimized,
and the global {\small MPI} communications are cut down to the minimum: for passing particles at the beginning of each time step,
and  for computing the long range FFTs.

@@

The paper is structured as follow: section \ref{sec:structure} reviews the structure and flow of the main code;
section \ref{sec:Poisson} describes how Poisson equation is solved on the two-mesh system;
we then present in section \ref{sec:scaling} the scaling of the code to very large problems,
including much larger runs that were produced at various high performance computing centres;
section \ref{sec:accuracy}  discuss the accuracy and systematic effects of the code.
 We then describe in section \ref{sec:halo} the run-time halo finder, in section \ref{sec:extensions} various extensions
 to the default configuration, and conclude afterwards.

@@

When it comes to finding haloes at run time, this buffer can create a problem, because a large object located close to the boundary can have a radius larger than the buffer zone, in which case it would be truncated and be assigned a wrong mass, centre of mass, etc. 
It could then be desirable to increase the buffer zone around each tile, at the cost of a loss of memory dedicated to the actual physical volume, and the code is designed to allow for such a flexibility.

@@

Because the coarse grid arrays require $4^3$ times less memory per node, 
it does not contribute much to the total memory requirement, and the bulk of the foot-print is 
concentrated in a handful of fine grid arrays.
Some of these are only required for intermediate steps of the calculations, 
hence it is possible to hide therein many coarse grid arrays\footnote{ This memory recycling is done with `equivalence' statements in {\small F90}}.   
We present here the largest arrays used by the code:
\begin{enumerate}
\item{{\tt xv} stores the position and velocity of each particle} 
\item{{\tt ll} stores the linked-list that accelerate the access to particles in each coarse grid cell}
\item{{\tt send\_buf, recv\_buf} store the particles to be {\small MPI}-passed to the neighbouring sub-volumes}
\item{{\tt rho\_f} and {\tt cmplx\_rho\_f} store 
the local fine grid density  in real and Fourier space respectively}
\item{{\tt force\_f} stores the force of gravity (short range only) along the three Cartesian directions}
\item{{\tt kern\_f} stores the fine grid force kernel in the three directions}
\item{{\tt PID} stores the unique particle identification tags.}
\end{enumerate}
The particle ID is a feature that can be switched off by removing a compilation flag, 
and allows to optimize the code for higher resolution configurations.

{\bf In terms of memory, {\small CUBEP$^3$M} can be configured such as the main footprint is dominated exclusively by the 
particle phase space array, the linked list and the buffer exchange arrays, in which case it asymptotes to about $44$ bytes per particles. 
For a fixed problem size, this can be achieved by maximizing the number of tiles, thereby reducing the size of the local fine mesh arrays.
There is a limit at which we can do such a volume breakdown, though, since the physical 
volume on each tile must be at least twice as large as the buffer, i.e. $\ge 96^3$ cells (see section \ref{sec:Poisson}).
In addition, this allows a high rate of particle exchange across nodes;  the memory per particle can be brought down to $32$ bytes 
if little particle passing is expected. 
However, minimizing the memory does not optimize the speed, as many {\small CPU}-intensive operations are performed on each tile -- {\small FFTW} for instance.
A higher number of cells per tile with fewer tiles usually run faster, in which case the other arrays in the above-mentioned list become memory heavy.
%
%uses about $130 - 150$ bytes per particles, and is in that sense more heavy than other codes like the 
For comparison, Lean-{\small GADGET-2} code uses  $84$ per particles \citep{Gadget2Lean}.}

%\subsection{Code overview}
%\label{subsec:overview}

@@

If the current redshift corresponds to one of the checkpoints, the code advances all particles to their final location
and writes them to file. 
These restart files start with a small header that 
contains the local  number of particles, the current redshift and time step, and the constraints on the time step jump
from the previous iteration; the header is followed by the position and velocity of each particle. The particle identification tags are written 
with a similar fashion in distinct files to simplify the post-processing coding. 
This general I/O strategy allows for highly efficient {\small MPI} parallel read and write,
by default in binary format for compactness, and has been shown to scale well to large data sets. 
At this stage, the performance depends only on the setup and efficiency of the parallel file system.

@@

Other constraints that need to be considered is that any decomposition geometry 
limits in some ways permissible grid sizes, and that volume evenly decomposed into cubic sub-sections 
--  such as {\small CUBEP$^3$M} -- requires the number of {\small MPI} processes to be a perfect cube.
Also,  since the decomposition is volumetric, as opposed to density dependent -- 
it suffers from load imbalance whenever the particles are not evenly distributed.
However, large scale cosmology problems are very weakly affected by this effect, which
can generally  be overcome by allocating a small amount of extra memory per node.
As mentioned in section \ref{sec:scaling}, large runs can even be optimized by  allocating 
less memory to start with, follow by a restart with a configuration that allows more memory per node, if required.

@@

The force of gravity on a mesh can be computed either with a gravitational potential  kernel $\omega_{\phi} ({\bf x})$
  or a force  kernel $\omega_{F} ({\bf x})$.
Gravity fields are curl-free, which allows us to relate the potential $\phi({\bf x})$ to the source term via Poisson's equation: 
\begin{eqnarray}
\nabla^{2}\phi({\bf x}) = 4 \pi G \rho({\bf x})
\label{eq:poisson}
\end{eqnarray}
We solve this equation in Fourier space, where we write:
\begin{eqnarray}
 \tilde{\phi}({\bf k}) = \frac{-4 \pi G \tilde{\rho}({\bf k})}{k^{2}} \equiv \tilde{\omega}_{\phi}({\bf k})\tilde{\rho}({\bf k})
\label{eq:poissonFourier}
\end{eqnarray}
The potential in real space is then obtained with an inverse Fourier transform, and the kernel becomes $\omega_{\phi} ({\bf x}) = -G/r$.
Using the convolution theorem, we can write
\begin{eqnarray}
 \phi({\bf x}) = \int \rho({\bf x'}) \omega_{\phi}({\bf x'} - {\bf x}) d{\bf x'}   
\label{eq:poisson_solution_pot}
\end{eqnarray}
and
\begin{eqnarray}
{\bf F}({\bf x}) = - m {\bf \nabla} \phi({\bf x}) 
\label{eq:Force_sol}
\end{eqnarray}
Although this approach is fast, it involves a finite differentiation at the final step, which enhances the numerical noise.
We therefore opt for a force kernel, which is more accurate, even though it requires four extra Fourier transforms.
In this case, we must solve the convolution in three dimensions and define the force kernel {\boldmath $\omega$}$_{F}$ such as:
\begin{eqnarray}
 {\bf F}({\bf x}) =  \int \rho({\bf x'}) \mbox{\boldmath $\omega$}_{F}({\bf x'} - {\bf x}) d{\bf x'}                                      
\label{eq:poisson_solution_force}
\end{eqnarray}
Because the gradient acting on Eq. \ref{eq:poisson_solution_pot} affects only unprime variables, we can express the force kernel as a gradient of the potential kernel. Namely:  
\begin{eqnarray}
\mbox{\boldmath $\omega$}_{F}({\bf x}) \equiv - {\bf \nabla}\omega_{\phi}({\bf x}) = - \frac{mG \hat{\bf r}}{r^{2}}
\end{eqnarray}

@@

Since these calculations are performed on two grids of different resolution, a sampling window function must be convoluted 
both with the density and the kernel (see [Eq. 7-8] of MPT).
When matching the two force kernels, the long range force is always on the low side close to the cutoff region, whereas the short range force is uniformly scattered across the theoretical $1/r^2$ value -- intrinsic features of the CIC and NGP interpolation schemes respectively.  By performing force measurements on two particles randomly placed in the volume, we identified a small region surrounding the cutoff length in which we empirically adjust both kernels such as to improve the match. Namely, for $14 \le r \le 16$, $\mbox{\boldmath $\omega$}_{s}({r}) \rightarrow 0.985\mbox{\boldmath $\omega$}_{s}({ r})$,
and for  $12 \le r \le 16$, $\mbox{\boldmath $\omega$}_{l}({r}) \rightarrow 1.2\mbox{\boldmath $\omega$}_{l}({ r})$.
{\bf A recent adjustment of the similar kind was also applied on the fine kernel to compensate for a systematic underestimate of the force at grid distances.
This is caused by the uneven balance of the force about the $1/r^2$ law in the NGP interpolation scheme.
Although most figures in this paper were made prior this correction, the new code configuration applies a 90 per cent boost to 
the kernel for the six elements that are adjacent, i.e. $\mbox{\boldmath $\omega$}_{s}({r=1}) \rightarrow 1.9\mbox{\boldmath $\omega$}_{s}({r=1})$ (see section \ref{subsec:force} for more details).}

@@

The choice of softening length is motivated by a trade off between accuracy and run time.
 Larger values reduce the structure formation but run quicker. 
 We show in Fig. \ref{fig:rsoft} the impact of changing this parameters on the power spectrum.
 For this test, we produced SciNet256 simulations, starting at $z=100$ and evolving to $z=0.5$, with a box size of $100 h^{-1}\mbox{Mpc}$.
 We also record in Table \ref{table:rsoft} the real run time for each trials, from where 
we see the strong effect of computation time the softening length has.
In this test, which is purposefully probing rather deep in the non-linear regime, reducing the softening length
to half its value ($1/20^{th}$ of a grid cell) double the run time. The gain in power spectrum is less than two per cent.
Similarly, increasing to $0.2$ the softening length reduces almost by a half the run time, but suffers from a five per cent loss in power at the resolution turn-around.
One tenth of grid cell seems to be the optimal choice in this trade off, and is therefore the default value, however it should really be considered a free parameter.

@@

\begin{table}
\begin{center}
\caption{Scaling in {\small CPU} resources as a function of the softening length, in units of fine grid cells.}
\begin{tabular}{|l|c|c|}
\hline 
$r_{soft}$      & time (h)   \\                 
\hline
 $0.5$ & 5.75 \\
 $0.3$ & 8.17\\
 $0.2$ & 10.09 \\
 $0.1$ & 18.67\\
 $0.05$ & 31.07 \\
% $0.03$ & XXX\\
% $0.01$ & XXX\\
\hline
\end{tabular}
\label{table:rsoft}
\end{center}
\end{table}

@@

We have also ran {\small CUBEP$^3$M} on a much larger number of cores, 
from 8000 to up to 21,976, with $5488^3$-$6000^3$ (165 to 216 billion) 
particles on Ranger and on JUROPA at the J\"ulich Supercomputing Centre in Germany, 
which is an Intel Xeon X5570 
(Nehalem-EP) quad-core 2.93 GHz system, also interconnected with Infiniband.
Since it is not practical to perform dedicated scaling tests on such a large number of
computing cores, we instead list in Table \ref{summary_scaling_table2} 
the data directly extracted from production runs. We have found the 
code to scale within 1.5 per cent of ideal up to 4096 cores. 
For larger sizes ($\ge$10,976 cores), the scaling is less ideal,
due to increased communication 
costs, I/O overheads (a single time slice of $5488^3$ particles is 3.6 TB)
and load balancing issues, but still within $\sim20$ per cent of ideal. 
These first three Ranger runs were performed 
with 4 {\small MPI} processes and 4 threads per Ranger node (`4way')\footnote{For these very large runs, 
we used a NUMA script {\it tacc\_affinity}, specially-provided by the technical staff, 
that bind the memory usage to local sockets, thus ensuring memory affinity. 
This becomes important because the memory sockets per node 
(32 GB RAM/node on Ranger) are actually not equal-access. Generally, the local 
memory of each processor has much shorter access time.}.

Furthermore, due to the increasing clustering of structures at those small 
scales, some of the cuboid sub-domains came to contain a number of particles
well above the 
average, thereby requiring more memory per {\small MPI} process
in order to run until the end. 
As a consequence,  throughout most of their late evolution, the 
largest two of these simulations were run with 4096 and 21,952 cores and 
with only 2 {\small MPI} processes and 8 threads per Ranger node (`2way'), which on 
Ranger allows using up to 16 GB of RAM per {\small MPI} process\footnote{In order to ensure 
local memory affinity,  a second special NUMA control script, {\it tacc\_affinity\_2way}, 
was developed for us by the TACC technical staff and allowed to run more efficiently 
in this mode.}. Because each processor accesses memory that is not fully local, this configuration does affect the 
performance somewhat, as does the imperfect load balancing that arises in such situations.
This can be seen in the rightmost point of Fig. \ref{scaling} (right panel), where the scaling is $42$ per cent below the ideal.
We note that we still get $\sim1.5$ speedup
from doubling the core count, even given these issues. Overall the
code scaling performance is thus satisfactory even at extremely large number of cores.
 We expect the code to handle even larger problems efficiently, and is thus well suited to run on
  next generation Petascale systems.

@@

 This section describes the systematics effects that are inherent to our P$^{3}$M algorithm.
 We start with a demonstration of the accuracy with a power spectrum measurement of a RANGER4000 simulation. 
 The halo mass function also assess the code capacity to model gravitational collapse, 
 but depends on the particular halo finder used.
 We thus postpone the discussion on this aspect until section \ref{sec:halo},
and focus for the moment on the particles only. In addition to the power spectrum, we further quantify the accuracy of the force calculation
 with comparisons to Newton's law of gravitation.

@@

\begin{figure*}%[ht]
  \begin{center}
%    \includegraphics[width=5.2in]{graphs/0.000xz_w_halos_pink_blue_enh-1.eps}
\includegraphics[width=5.2in]{graphs/0.000proj_1000x1000.eps}
  \caption{Detail of a dark matter density $3.2 h^{-1}\mbox{Gpc}$Êper side, projected over $15 h^{-1}\mbox{Mpc}$ and measured at $z=0$ in a RANGER4000 simulation.
  Particles are shown in pink, haloes in blue in the online version, and the colour scale is linear. The full projection is 64 times larger,
  but we show here only a sub-volume, which provides more details on the structure.
    \label{fig:density}}
\end{center}
\end{figure*}

@@

The top right panel of Fig. \ref{fig:den_force_fracErr} shows the fractional error on the force along the radial direction (top) and the fractional 
tangential contribution (bottom), also calculated from a single time step.
Again, particles that are in the same cell have zero fractional error in the radial direction, and zero tangential force.
Particles feel a large scatter from the neighbouring fine mesh cells, but as the distance increase beyond five fine cells, the fractional error drops down to about 10 per cent.
%The fractional error of the radial component is up to 60 per cent at sub-grid distances, but remains under ten per cent for  distances larger than a few cells. 
The transverse fractional error peaks exactly at the grid distance,  and is about 50 per cent smaller than the radial counterpart.

Although these scatter plots show the contributions from individual particles and cells, it is not clear
whether the mean radial force felt is higher or lower the predictions, hence we rebin the results
in 50 logarithmically spaced bins and compute the mean and standard deviation. 
We show this plot in the middle left panel of Fig. \ref{fig:den_force_fracErr}, where we observe that on average,
there is a 50 per cent fractional error in the range $ 0.4 < r < 2$ (in units of fine grid), but the numerical calculations are otherwise
in excellent agreement with Newton's law.
Since  the transverse force is, by definition, a positive number,
and since we know that it averages out to zero over many time steps, we plot only the scatter about the mean, represented in the figure by the solid line.
